---
title: "Bulgaria's Weather & Climate"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(terra)
library(stars)
library(tidyverse)
library(lubridate)
library(rgdal)
library(sf)
library(ggthemes)
library(here)
```

Project 0: Compare daily observational data with data from downscaling projects.

Project 1: Create a space-time object of daily high/low temperatures and precipitation for all 1x1 km grid cells inside the boundary defined by Sofia grad over the period 1950-2020. Then examine climatologies / trends at specific locations (e.g., apartment).

Sofia Bulgaria weather and climate station (WMO 15614). https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:BUM00015614/detail
42.695°, 23.406° Temperatures in F and precipitation in inches (hundreths). 00Z to 00Z
```{r}
SOF.df <- read.csv(file = "SOF_1952_2021.csv") |>
  mutate(Date = as.Date(DATE),
         Prcp = PRCP * 25.4) |>
  select(Date, Tmax = TMAX, Tmin = TMIN, Tavg = TAVG, Prcp)
```


Check out the {climate} package for downloading station observation data. https://www.rdocumentation.org/packages/climate/versions/1.0.3
```{r}
remotes::install_github("bczernecki/climate")
library(climate)
```

Find nearest stations.
```{r}
nearest_stations_ogimet(country = "Bulgaria",
  date = Sys.Date(),
  add_map = TRUE,
  point = c(23.406, 42.695),
  no_of_stations = 100
)

# 15614 Sofia Observ. 23.38334 42.65001  586 meters lon/lat does not match

( SOF2.df <- meteo_ogimet(date = c(as.Date("2020-02-20"), as.Date("2020-02-22")), 
                        interval = "daily",
                        coords = FALSE, 
                        station = 15614) |>
  mutate(Tmax = TemperatureCMax * 1.8 + 32,
         Tmin = TemperatureCMin * 1.8 + 32,
         Tavg = TemperatureCAvg * 1.8 + 32,
         Prcp = Precmm) |>
  select(Date, Tmax, Tmin, Tavg, Prcp) |>
  arrange(Date) )
```

Daily are 6am to 6am (9 UTC to 9 UTC) and put on the second morning date.

```{r}
Z <- SOF2.df |>
  left_join(SOF.df, by = "Date")
```

Turn 3-hourlies into daily max/min, precip.
```{r}
( SOF3.df <- meteo_ogimet(date = c(as.Date("2020-01-01"), as.Date("2020-07-08")), 
                          interval = "hourly",
                          coords = FALSE, 
                          station = 15614) |>
  select(Date, TC, TmaxC, TminC, pr6, pr12, pr24) |>
  arrange(Date) )

( SOF4.df <- SOF3.df |>
  mutate(Date = with_tz(Date, "EET"),
         YMD = as.Date(Date)) |>
  group_by(YMD) |>
  summarize(Tmax_C = max(as.numeric(TmaxC), na.rm = TRUE),
            Tmin_C = min(as.numeric(TminC), na.rm = TRUE),
            Prcp_mm = sum(pr12, na.rm = TRUE)) )
```

```{r}
Z <- SOF4.df |>
  mutate(Date = YMD) |>
  left_join(SOF.df, by = "Date")
```


Administrative boundaries https://data.humdata.org/dataset/bulgaria-administrative-level-0-2-boundaries
```{r}
#unzip(zipfile = "bgr_adm_unicef_2019_shp.zip", 
#      exdir = "Bulgaria")
Bulgaria0.sf <- st_read(dsn = "Bulgaria",
                        layer = "bgr_admbnda_adm0_UNICEF_2019")
plot(Bulgaria0.sf$geometry)

Bulgaria1.sf <- st_read(dsn = "Bulgaria",
                        layer = "bgr_admbnda_adm1_UNICEF_2019")
plot(Bulgaria1.sf$geometry)

SofiaCity.sf <- Bulgaria1.sf %>%
  filter(ADM1_EN == "Sofia-City")

Bulgaria2.sf <- st_read(dsn = "Bulgaria",
                        layer = "bgr_admbnda_adm2_UNICEF_2019")
plot(Bulgaria2.sf$geometry)

library(crsuggest)
suggest_crs(Bulgaria0.sf)
```

Also the European Environment Agency (EEA) reference grid shapefiles at 1, 10, and 100 km resolution from https://www.eea.europa.eu/data-and-maps/data/eea-reference-grids-2/gis-files/bulgaria-shapefile
```{r, eval=FALSE}
#unzip(zipfile = "Bulgaria_shapefile.zip", 
#      exdir = "Bulgaria")
Bulgaria.sf <- st_read(dsn = "Bulgaria",
                       layer = "bg_10km")
plot(Bulgaria.sf$geometry)
st_crs(Bulgaria.sf)
```

Get European daily precip and tmax, tmin data from a cloud-optimized geotiff (COG) hosted at ftp://palantir.boku.ac.at/Public/ClimateData/ See: Readme.txt

Accessing data from cloud optimised geotiffs with R: https://frodriguezsanchez.net/post/accessing-data-from-large-online-rasters-with-cloud-optimized-geotiff-gdal-and-terra-r-package/

Temperatures are given as celsius * 100 and precipitation is given as mm * 100. For questions please contact: mathias.neumann@boku.ac.at v3 1950-2020. Citation: Moreno, A., & Hasenauer, H. (2015). Spatial downscaling of European climate data. International Journal of Climatology.DOI: 10.1002/joc.4436 https://rmets.onlinelibrary.wiley.com/doi/10.1002/joc.4436

Precipitation rasters. About 12 secs per year. Start by making SofiaCity.v a vect object.
```{r}
SofiaCity.v <- vect(SofiaCity.sf)

t0 <- Sys.time()

cog.url0 <- "/vsicurl/ftp://palantir.boku.ac.at/Public/ClimateData/v3_cogeo/AllDataRasters/prec/DownscaledPrcp2018_cogeo.tif"
cog.url1 <- "/vsicurl/ftp://palantir.boku.ac.at/Public/ClimateData/v3_cogeo/AllDataRasters/prec/DownscaledPrcp2019_cogeo.tif"
cog.url2 <- "/vsicurl/ftp://palantir.boku.ac.at/Public/ClimateData/v3_cogeo/AllDataRasters/prec/DownscaledPrcp2020_cogeo.tif"

urls <- c(cog.url0, cog.url1, cog.url2)

ppt.r <- rast(urls) |>
  terra::crop(SofiaCity.v)

Sys.time() - t0
```

Convert to a {stars} object.
```{r}
ppt.st <- st_as_stars(ppt.r)

dates <- seq.Date(from = as.Date("2018-01-01"),
                  to = as.Date("2020-12-31"),
                  by = "days")

ppt.st <- ppt.st |>
  st_set_dimensions(which = "band", 
                    values = dates, 
                    names = "time") / 100

ppt.st <- ppt.st |>
  st_crop(SofiaCity.sf) |>
  setNames("Precipitation_mm")
```

Check precipitation values over 30 days in July.
```{r}
X <- filter(ppt.st, 
            time >= ymd("2018-07-01"), 
            time <= ymd("2018-07-30"))

ggplot() +  
  geom_stars(data = X["Precipitation_mm"], alpha = .8, downsample = c(0, 0, 0)) + 
  facet_wrap(~ time) +
  scale_fill_viridis_c() +
#  scale_fill_distiller(palette = "RdBu") +
  coord_equal() +
  theme_map() +
  theme(legend.position = "bottom") +
  theme(legend.key.width = unit(2, "cm"))
```

The down-scaling smooths spatial variations as seen by the fact that if at least one cell gets rain there is rain over most of the city. 

Compare with daily rainfall rasters from CHIRPS (climate hazards infrared precipitation with stations). https://www.chc.ucsb.edu/data/chirps. https://www.nature.com/articles/sdata201566
```{r}
# remotes::install_github("mikejohnson51/AOI")
# remotes::install_github("mikejohnson51/climateR")
library(raster)
library(AOI)
library(climateR)
select <- dplyr::select  # make sure that the select you want is from {dplyr} (not the one in {raster})

country_of_interest <- "Bulgaria"
AOI <- aoi_get(country = country_of_interest)
plot(AOI$geometry)
```

Daily rainfall in Sofia city.
```{r}
t0 <- Sys.time()
start <- "2018-07-01"
end <- "2018-07-30"
ppt.rs <- climateR::getCHIRPS(SofiaCity.sf, 
                              startDate = start, 
                              endDate = end) 
Sys.time() - t0

dates2 <- seq.Date(from = as.Date(start),
                   to = as.Date(end),
                   by = "days")
ppt2.st <- st_as_stars(ppt.rs) |>
  st_set_dimensions(which = "band", 
                    values = dates2, 
                    names = "time") |>
  st_crop(SofiaCity.sf, 
          crop = TRUE) |>
  setNames("Precipitation_mm")

ggplot() +  
  geom_stars(data = ppt2.st["Precipitation_mm"], alpha = .8, downsample = c(0, 0, 0)) + 
  facet_wrap(~ time) +
  scale_fill_viridis_c() +
#  scale_fill_distiller(palette = "RdBu") +
  coord_equal() +
  theme_map() +
  theme(legend.position = "bottom") +
  theme(legend.key.width = unit(2, "cm"))

ppt2.st |> 
  as_tibble() |>
  group_by(time) |> 
  summarize(avgR = mean(Precipitation_mm, na.rm = TRUE))
```






Daily rainfall amounts in Bulgaria March 2021.
```{r}
start <- "2021-03-01"
end <- "2021-03-31"
df <- climateR::getCHIRPS(Bulgaria0.sf, 
                          startDate = start, 
                          endDate = end) 

prcp.st <- st_as_stars(df)
names(prcp.st) <- "mm"

days <- seq(as_date(start), 
            as_date(end), 
            by = "day")

( prcp.st <- st_set_dimensions(prcp.st, 
                               which = 3, 
                               values = days, 
                               names = "date") )

X <- st_crop(prcp.st, 
             Bulgaria0.sf, 
             crop = TRUE)

ggplot() +  
  geom_stars(data = X) + 
  facet_wrap("date") +
  scale_fill_viridis_c() +
  coord_sf() +
  theme_map() +
  theme(legend.position = "bottom") +
  theme(legend.key.width = unit(2, "cm"))
```

Extract values at a particular location.
```{r}
# center lat/lon Sofia City district (23.39864 42.67827)
#coords.center <- data.frame(lon = 23.39864, lat = 42.67827)

coords.apartm <- data.frame(lon = 23.36161, lat = 42.67856)
precip_apartm <- terra::extract(ppt.r,
                                coords.apartm,
                                list = TRUE) |>
  unlist()
```

Daily high temperature in 2010
```{r}
cog.url <- "/vsicurl/ftp://palantir.boku.ac.at/Public/ClimateData/v3_cogeo/AllDataRasters/tmax/DownscaledTmax2020_cogeo.tif"
tmax.r <- rast(cog.url)
tmax.r
```

1 km elevation raster over Europe. Values are not elevation. https://www.eea.europa.eu/data-and-maps/data/digital-elevation-model-of-europe
```{r}
elev.r <- rast("elevation1x1_new.tif")
plot(elev.r)

elev.r <- terra::project(elev.r, ppt.r)
```

Convert daily `SpatRaster` object to a data frame with each row a grid location and columns 3:365 the daily values.
```{r}
Tmax.df <- as.data.frame(Tmax.r, xy = TRUE) %>%
  set_names(nm = c("lon", "lat", str_c("D", 1:365)))
head(Tmax.df[, 1:10])
range(Tmax.df[,-(1:2)])
```






The database of Open Street Maps https://dominicroye.github.io/en/2018/accessing-openstreetmap-data-with-r/
```{r}
library(osmdata)
library(ggmap)

head(available_features())
head(available_tags("amenity"))
head(available_tags("shop"))
```

Build a query.
```{r}
q <- getbb("Sofia, Bulgaria") %>%
  opq() %>%
  add_osm_feature(key = "surface")
str(q)

surfaceRoads <- osmdata_sf(q)
plot(surfaceRoads$osm_lines$geometry)

q <- getbb("Sofia, Bulgaria") %>%
  opq() %>%
  add_osm_feature(key = "public_transport")
str(q)

publicTransport <- osmdata_sf(q)
plot(publicTransport$osm_lines$geometry)

q <- getbb("Sofia, Bulgaria") %>%
  opq() %>%
  add_osm_feature(key = "amenity",
                  value = "restaurant")
str(q)

restaurants <- osmdata_sf(q)
```

Make a map.
```{r}
mad_map <- get_map(getbb("Sofia, Bulgaria"), 
                   maptype = "toner-background",
                   source = "stamen")

coords.apt.sf <- st_as_sf(coords.apt.v) 
st_crs(coords.apt.sf) <- 4326

ggmap(mad_map) +
  geom_sf(data = coords.apt.sf,
          inherit.aes = FALSE,
          colour = "red")
```



Read CHIRPS with {stars} https://github.com/loreabad6/chirps-stars
```{r}
url <- "https://data.chc.ucsb.edu/products/CHIRPS-2.0/global_daily/netcdf/p05/"
file <- "chirps-v2.0.1981.days_p05.nc"
fn <- paste0(url, file)

fn <- "chirps-v2.0.1981.01.01.tif"

chirps1981 <- read_stars(fn, NA_value = -9999)
chirps1981

url <- "https://data.chc.ucsb.edu/products/CHIRPS-2.0/global_daily/tifs/p05/1981/"
file <- "chirps-v2.0.1981.01.01.tif.gz"

download.file(url = paste0(url, file),
              destfile = "data/temp.zip.gz")
library(R.utils)
gunzip("data/temp.zip.gz")

chirps1981 <- read_stars("data/temp.zip.gz", NA_value = -9999)
```

Get climate data: 1958-2019
```{r}
df <- climateR::getTerraClim(Bulgaria0.sf, 
                             param = "tmax", 
                             startDate = "2019-07-04", 
                             endDate = "2019-07-04")[[1]]

plot(df)
plot(Bulgaria0.sf$geometry, 
     add = TRUE)

SofiaCity.sf <- Bulgaria1.sf %>%
  dplyr::filter(ADM1_EN == "Sofia-City")

df <- climateR::getTerraClim(SofiaCity.sf, 
                             param = "tmax", 
                             startDate = "2019-07-01", # monthly only
                             endDate = "2019-07-01")
plot(df$terraclim_tmax)
plot(SofiaCity.sf$geometry, 
     add = TRUE)
```

Convert to {stars} object.
```{r}
library(stars)

df <- climateR::getTerraClim(SofiaCity.sf, 
                             param = "tmax", 
                             startDate = "2019-01-01", 
                             endDate = "2019-12-01")

tmax.st <- st_as_stars(df$terraclim_tmax)

names(tmax.st) <- "C"

months <- seq(as_date("2019-01-01"), 
              as_date("2019-12-01"), 
              by = "month")

months <- month.abb

( tmax.st <- st_set_dimensions(tmax.st, 
                               which = 3, 
                               values = months, names = "month") )

X <- st_crop(tmax.st, 
             SofiaCity.sf, 
             crop = TRUE)

ggplot() +  
  geom_stars(data = X) + 
  facet_wrap("month") +
  scale_fill_viridis_c() +
  coord_sf() +
  theme_map() +
  theme(legend.position = "bottom") +
  theme(legend.key.width = unit(2, "cm"))
```




